---
title: "ARIMA modeling Techniques to Real-World Time Series."
author: "Stephen"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r}
# Load required packages
library(forecast)
library(tseries)

# Load the AirPassengers dataset
data(AirPassengers)
ts_data <- AirPassengers

# Step 1: Plot the time series and describe its pattern
plot(ts_data, main="AirPassengers Time Series", ylab="Passenger Numbers", xlab="Year")

```
Description of General Pattern (Step 1):
	•	The AirPassengers dataset shows monthly international airline passengers from 1949 to 1960.
	•	Trend: There is a clear upward trend, indicating passenger numbers increase over time.
	•	Seasonality: A repeating pattern within each year suggests strong seasonality (likely 12-month periodicity due to monthly data).
	•	Variance: The amplitude of seasonal fluctuations increases over time, suggesting non-constant variance.

```{r}
# Step 2: Apply logarithmic transformation to stabilize variance
log_ts_data <- log(ts_data)
plot(log_ts_data, main="Log-Transformed AirPassengers", ylab="Log(Passenger Numbers)", xlab="Year")

```
Comparison (Step 2):
	•	The log transformation reduces the increasing variance observed in the original series.
	•	The seasonal fluctuations appear more consistent in amplitude, and the trend remains upward but is now more linear.

```{r}
# Step 3: Apply differencing to achieve stationarity
# First differencing
diff_log_ts <- diff(log_ts_data, differences=1)
plot(diff_log_ts, main="First Differenced Log AirPassengers", ylab="Differenced Log(Passengers)", xlab="Year")

# Seasonal differencing (period = 12 for monthly data)
diff_log_ts_seas <- diff(diff_log_ts, lag=12)
plot(diff_log_ts_seas, main="First + Seasonal Differenced Log AirPassengers", ylab="Differenced Log(Passengers)", xlab="Year")

# Test for stationarity using ADF test
adf.test(diff_log_ts_seas, alternative="stationary")

```
Interpretation (Step 3):
	•	First Differencing: Removes the trend, but some seasonality remains.
	•	Seasonal Differencing: Applying a lag-12 difference (since the data is monthly) removes the seasonal pattern.
	•	The Augmented Dickey-Fuller (ADF) test on the differenced series typically yields a low p-value (< 0.05), indicating stationarity.

```{r}

# Step 4: Plot ACF and PACF of the transformed and differenced data
par(mfrow=c(1,2))
acf(diff_log_ts_seas, main="ACF of Differenced Series")
pacf(diff_log_ts_seas, main="PACF of Differenced Series")

```

Interpretation (Step 4):
	•	ACF: Shows significant spikes at lag 1 and possibly at seasonal lags (e.g., 12, 24), indicating potential MA components.
	•	PACF: Shows significant spikes at lag 1 and possibly at seasonal lags, suggesting potential AR components.
	•	The patterns suggest a seasonal ARIMA model with non-seasonal and seasonal components.

Step 5: Propose Two Candidate ARIMA Models Based on the ACF and PACF:
	•	Model 1: ARIMA(0,1,1)(0,1,1)[12]
	◦	Non-seasonal: MA(1) due to significant ACF spike at lag 1.
	◦	Seasonal: MA(1) with period 12 due to seasonal ACF spike.
	•	Model 2: ARIMA(1,1,1)(1,1,1)[12]
	◦	Non-seasonal: AR(1) and MA(1) due to PACF and ACF spikes.
	◦	Seasonal: AR(1) and MA(1) for seasonal components.
	
```{r}
# Step 6: Fit the proposed ARIMA models
model1 <- Arima(log_ts_data, order=c(0,1,1), seasonal=c(0,1,1))
model2 <- Arima(log_ts_data, order=c(1,1,1), seasonal=c(1,1,1))

# Use auto.arima for comparison
auto_model <- auto.arima(log_ts_data)

```

```{r}
# Step 7: Compare models using AIC/BIC
summary(model1)$aic
summary(model1)$bic
summary(model2)$aic
summary(model2)$bic
summary(auto_model)$aic
summary(auto_model)$bic

```
Comparison (Step 7):
	•	Lower AIC/BIC values indicate a better model fit.
	•	Typically, auto.arima may select a model like ARIMA(0,1,1)(0,1,1)[12], but we compare all three.
	
	•	output:
	◦	Model 1: AIC = -483.3991, BIC = -474.7735
	◦	Model 2: AIC = -480.3109, BIC = -465.9349
	◦	Auto Model: AIC = -483.3991, BIC = -474.7735
	•	We choose the model with the lowest AIC/BIC (auto.arima or Model 1).

```{r}
# Step 8: Extract and interpret parameters of the best model
best_model <- model1  # Model 1 has the lowest AIC/BIC
coef(best_model)
summary( best_model)
```

Interpretation (Step 8):
• For ARIMA(0,1,1)(0,1,1)[12]:
◦ ma1: The non-seasonal MA(1) coefficient is -0.4018, indicating a moderate negative adjustment where the current value is influenced inversely by the previous error term (e.g., overcorrections for positive shocks in the series).
◦ sma1: The seasonal MA(1) coefficient is -0.5569, reflecting a stronger negative seasonal adjustment that captures recurring error patterns across the 12-month cycle.
• Both coefficients are statistically significant, as their absolute values exceed twice their standard errors (s.e. of 0.0896 for ma1 and 0.0731 for sma1), implying reliable estimates.
• The model's residual variance (sigma²) is low at 0.001371, with a log likelihood of 244.7, AIC of -483.21, and BIC of -474.77, all supporting a strong overall fit.
• Training set error measures further validate this: RMSE = 0.0350 (low forecast error), MAE = 0.0263 (average absolute deviation), and MAPE = 0.475% (high percentage accuracy), with minimal bias (ME ≈ 0.0006) and no notable autocorrelation in residuals (ACF1 ≈ 0.014).

```{r}
# Step 9: Plot residuals and their ACF
residuals <- residuals(best_model)
plot(residuals, main="Residuals of Best Model")
acf(residuals, main="ACF of Residuals")

```

Interpretation (Step 9):
• The residual plot displays fluctuations centered around zero, with values ranging from approximately -0.10 to 0.10 across the time period (1949–1960). There is no discernible trend, seasonality, or heteroscedasticity, as the amplitude remains relatively constant, closely resembling white noise.
• The ACF plot of the residuals shows all autocorrelation coefficients well within the 95% confidence bands (dashed blue lines), with no spikes exceeding these thresholds at any lag. This confirms that the residuals are uncorrelated, indicating the model has effectively captured the underlying patterns in the data.

```{r}
# Step 10: Perform Ljung-Box test
Box.test(residuals, lag=24, type="Ljung-Box")

```
Interpretation (Step 10):
• The Ljung-Box test yields a test statistic of 26.446 with 24 degrees of freedom and a p-value of 0.3399, indicating no significant autocorrelation in the residuals.
• With a p-value > 0.05, the residuals are consistent with white noise, suggesting the ARIMA(0,1,1)(0,1,1)[12] model adequately captures the structure of the AirPassengers data.
• This high p-value (e.g., 0.3399) supports model adequacy by confirming that the residuals exhibit no systematic patterns or dependencies.


Interpretation (Step 11):
• The model is deemed adequate if the following conditions are met:
  ◦ Residuals exhibit no discernible patterns, resembling white noise.
  ◦ ACF of residuals shows no significant spikes beyond the confidence bands.
  ◦ Ljung-Box test p-value exceeds 0.05, indicating no significant autocorrelation.
• Based on the analysis, the ARIMA(0,1,1)(0,1,1)[12] model is adequate, as the residual plot shows random fluctuations without trend or seasonality, the ACF lacks significant spikes, and the Ljung-Box test yields a p-value of 0.3399 (> 0.05). Additionally, the model's low AIC (-483.21) and BIC (-474.77) further support its suitability for the AirPassengers dataset.

```{r}	
# Step 12 & 13: Forecast the next 12 months and plot
forecast <- forecast(best_model, h=12)
plot(forecast, main="12-Month Forecast for AirPassengers", ylab="Log(Passenger Numbers)")

```

Interpretation (Step 13):
	Interpretation (Step 13):
• The forecast illustrates predicted log-passenger numbers for the period extending into 1961, continuing the upward trend observed in the historical data from 1949 to 1960.
• The 95% confidence intervals, shown as shaded blue regions, widen progressively over the 12-month forecast horizon, reflecting increasing uncertainty as predictions extend further into the future. This is consistent with the compounding effect of forecast errors over time.


Step 14: Comment on Expected Passenger Numbers and Uncertainty
	•	Expected Numbers: Convert forecasts back to the original scale using exp(forecast$mean). For example, if the log-forecast is ~6.5, then passenger numbers are ~665 (exp(6.5)).
	•	Uncertainty: The confidence intervals (e.g., 600–730 passengers) widen, indicating higher uncertainty for later months due to the compounding effect of errors.
	
	
```{r}
# Step 15: Compare ARIMA with ETS model (using out-of-sample evaluation for better assessment)

# Split data (hold out last 12 months for test)
train <- window(log_ts_data, end = c(1959, 12))
test <- window(log_ts_data, start = 1960)

# Refit models on train
best_model <- Arima(train, order = c(0,1,1), seasonal = c(0,1,1))
ets_model <- ets(train)

# Forecast the test horizon
forecast <- forecast(best_model, h = 12)
ets_forecast <- forecast(ets_model, h = 12)

# Plot ETS forecast (on full data for visualization, but evaluation is out-of-sample)
ets_model_full <- ets(log_ts_data)  # For plotting on full data if desired
ets_forecast_full <- forecast(ets_model_full, h=12)
plot(ets_forecast_full, main="ETS Forecast for AirPassengers", ylab="Log(Passenger Numbers)")

# Compare out-of-sample accuracy
accuracy(forecast, test)
accuracy(ets_forecast, test)

```
Interpretation (Step 15):
• A detailed comparison of metrics derived from the `accuracy()` function reveals that the ARIMA(0,1,1)(0,1,1)[12] model outperforms the ETS model in out-of-sample prediction on the test set. Specifically, the ARIMA model achieves a Root Mean Squared Error (RMSE) of 0.22210642, compared to the ETS model's RMSE of 0.38514455. This 42% reduction in RMSE (from 0.385 to 0.222) highlights the ARIMA model's superior ability to minimize the average magnitude of forecast errors, making it more reliable for predicting the AirPassengers data beyond the training period (1959–1960).
• The Mean Absolute Error (MAE) further supports this, with ARIMA at 0.18391448 and ETS at 0.31671965, indicating that the ARIMA model's average absolute deviation from actual values is approximately 42% lower. Other metrics, such as Mean Percentage Error (MPE) and Mean Absolute Percentage Error (MAPE), also favor ARIMA (-0.37371965 vs. -0.567208, and 4.736254 vs. 7.574738, respectively), though these are less critical here due to the logarithmic scale of the data.
• The ARIMA model's strength lies in its tailored specification, which effectively captures the dataset's clear upward trend and 12-month seasonal pattern, as informed by the earlier differencing and ACF/PACF analysis. This alignment with the data's structure is reflected in its lower error metrics, particularly on the test set, where it generalizes better.
• Conversely, the ETS model, likely employing a Holt-Winters method, approaches trend and seasonality through a different smoothing framework. While it performs reasonably well in-sample (RMSE = 0.08042144), its out-of-sample RMSE of 0.38514455 suggests it struggles to adapt to the test data's dynamics, possibly due to over-sensitivity to historical patterns or less flexibility in handling the specific seasonal structure. This discrepancy underscores that ARIMA's parametric approach may be better suited to this dataset's clear and consistent temporal characteristics.

```